A note on observation processes in epidemic models

### 2019-07-30 ms/ms.tex
tweaks to writing

Note that incidence is different from prevalence,
$I(t)$, which is defined as the number of \emph{currently} infected individuals;
the observed dynamics of incidence depend on the reporting time step (because the sum of 
true incidence is equal to the final size of an epidemic), whereas those of
prevalence do not. We expect the dynamics of incidence and prevalence to be similar only when
the reporting time step is equal to the disease generation time. While it is
uncommon, some models do not make a clear distinction between the two (e.g., 
\cite{hooker2010parameterizing}).

@@

We assumed that newly infected cases are reported instantaneously when
individuals become infected or when they recover. Although neither of these
assumptions is realistic, they are broadly consistent with the assumptions
of other state-of-art epidemic models (see Introduction). 
More realistic models may distinguish reported and unreported
(identified and unidentified) cases by adding
new state variables \citep{browne2015modeling,webb2015model} 
or by modeling an explicit delay kernel in reporting time 
\citep{harris1990reporting, ferguson2001foot, ster2009epidemiological, 
birrell2011bayesian, funk2018real}.

While we have only considered observation processes associated with incidence 
reporting in this study, it is equally as important to consider observation
processes in mortality reporting. In particular, many disease modelers have
tried to infer underlying transmission mechanisms from historical mortality 
data but assumed that individual deaths are recorded as soon as individuals die 
\citep{he2013inferring, didelot2017model, dean2018human}. 
This assumption does not account for the possibility 
that a delay in reporting of deaths can
change the shape of an epidemic curve: longer delays are expected to decrease
the size of an epidemic peak and delay the timing of the peak (Figure 1). 
Ignoring potential delays in reporting may affect their conclusions about the underlying 
transmission mechanisms.

Our study shows that seemingly irrelevant changes in the assumptions of an epidemic
model affects the inference of infectious disease transmission.
We caution disease modelers to be more mindful about their decisions in developing
epidemic models and the implications of their model assumptions.

>>>>

Note that incidence is different from prevalence,
$I(t)$, which is defined as the number of \emph{currently} infected individuals;
the observed dynamics of incidence depend on the reporting time step (because the sum of 
true incidence is equal to the final size of an epidemic), whereas those of
prevalence do not. We expect the dynamics of incidence and prevalence to be similar only when
the reporting time step is equal to the disease generation time. While they are relatively
uncommon, some models do not make a clear distinction between a discrete-time observed 
variable and a continuous-time latent variable (Table 1).

\begin{table}
\centering
\begin{tabular}{l | c | c}
\hline
\textbf{Study} & \textbf{Disease} &  \textbf{Observed variables}\\
\hline
\cite{capistran2009parameter} & Respiratory syncytial virus & Number of hospital cases\\
\cite{hooker2010parameterizing} & Measles & Number of reported cases\\
\cite{eisenberg2013identifiability} & Cholera & Number of reported cases\\
\cite{yang2013stability} & Hand-foot-mouth disease & Number of reported cases\\
\cite{gonzalez2014fractional} & Influenza & Number of confirmed cases\\
\hline
\end{tabular}
\caption{
\textbf{List of studies that fit prevalence curves to incidence data.}
Incidence, defined as the number of new cases between two reporting periods, depends
on the reporting time step.
Prevalence, defined as the number of currently infected individuals, does not depend
on the reporting time step.
Some models do not make a clear distinction between these two quantities.
}
\end{table}

@@

We assumed that newly infected cases are reported instantaneously when
individuals become infected or when they recover. Although neither of these
assumptions is realistic, they are broadly consistent with the assumptions
of other state-of-art epidemic models (see Introduction). 
More realistic models may distinguish reported and unreported
(i.e., identified and unidentified) cases by adding
new state variables \citep{browne2015modeling,webb2015model} 
or by modeling an explicit delay kernel in reporting time 
\citep{harris1990reporting, ferguson2001foot, ster2009epidemiological, 
birrell2011bayesian, funk2018real}.

While we only considered observation processes associated with incidence 
reporting in this study, it is equally as important to consider observation
processes in mortality reporting. In particular, many disease modelers have
tried to infer underlying transmission mechanisms from historical mortality 
data but assumed that individual deaths are recorded as soon as individuals die 
\citep{he2013inferring, didelot2017model, dean2018human}; this includes
\cite{kermack1927contribution} who approximated the reported \emph{number} of
deaths per week from plague with instantaneous death \emph{rates} ($dR/dt$).
This assumption does not account for the possibility 
that a delay in reporting of deaths can
change the shape of an epidemic curve: longer delays are expected to decrease
the size of an epidemic peak and delay the timing of the peak (Figure 1). 
Ignoring delays in reporting may affect their conclusions about the underlying 
transmission mechanisms.

Our study shows that a seemingly negligible change in the assumptions of an epidemic
model can affect the inference of infectious disease transmission.
We caution disease modelers to be more mindful about their decisions in developing
epidemic models and the implications of their model assumptions.
